---
title: "Climate change impacts in Latin America"
Author: varios
format:
  html: default
  docx: 
     reference-doc: custom-reference-doc.docx
bibliography: references.bib
include-in-header:
  - text: |
     <link rel = "shortcut icon" href = "img/ai-generated-8438225_1280.png" />
engine: knitr
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup


pacman::p_load(terra, geodata)

```


## Introduction

Climate change in Latin America

Taking the Sixth Assessment Report by the IPCC [@lee2023] as an starting point, we look into the development and forecasts of climate change for Latin America and the Caribbean. This chapter explore the impacts of climate change over mammals from  Latin America and the Caribbean region. This part of the world is a intricate natural geographic span with a complex ecological and political structure. To be specific describing the region, we adhere to the United Nations M49 standard [@standard1999] developed to facilitate the consolidation of national statistical data to regional levels. According to M49 standard Latin America is a region formed by three subregions: Caribbean (28 islands, spanning 15 countries), Central America (8 countries), and South America (16 countries). This regions extends over 20,139,378 km<sup>2</sup> of land. This territory is a little more than 13% of world emerged land. Moving into the see the reference to the Exclusive Economic Zone results in a total area of 140,776,000 km<sup>2</sup> (almost 40% of ocean surface)


```{r}
mapa <- vect("~/1 Nubes/OneDrive - El Instituto de Ecología/America Latina y el Caribe/america_latina_caribe_eez_v12.gpkg")
plot(mapa, values = mapa$GEONAME, legend = F)

mapa2 <- vect("~/1 Nubes/OneDrive - El Instituto de Ecología/America Latina y el Caribe/Morrone_Provincias_Biogeograficas/America Latina y Caribe.shp")

#test1 <- terra::(mapa, mapa2)
#plot(test1)
```


 


```{r}
#| label: inicial
#| echo: false
#| message: false
#| warning: false


# ============================
# Mapas "2030" (2021–2040) vs 1970–2000 para América Latina
# Variables: Temperatura media anual (Δ °C) y Precipitación anual (Δ %)
# SSP se refiere a "Shared Socio-economic Pathways" (SSPs): 126, 245, 370 and 585. 
# The two last digits referto Representative Concentration Pathways (RCP), expressed
# in terms of the increased capacity to trap energy by the year 2100, measured im W/m²
#     SSP1 being the "Sustainability" with forcing level 2.6 W/m²
#     SSP2 "Middle of the Road"  (forcing 4.5 W/m²))
#     SSP3 "Regional Rivalry"  (3.7 W/m²)
#     SSP5 "Fossil-fueled Development" (8.5 W/m²)
# Fuentes: WorldClim v2.1 (baseline) y CMIP6 bias-corrected (futuro); Natural Earth (límites)
#          https://www.carbonbrief.org/cmip6-the-next-generation-of-climate-models-explained/
# ============================

# Bibliotecas
pacman::p_load("terra", "sf", "geodata", "rnaturalearth", 
               "rnaturalearthdata", "ggplot2", "dplyr", 
               "purrr", "stringr",
               "patchwork")


# Funciones auxiliares ----
to_df <- function(r, aoi_sf) 
  {
    # Recorta y enmascara al AOI, convierte a data.frame para ggplot
    r1 <- crop(r, vect(aoi_sf)) |> 
      mask(vect(aoi_sf))
    
      df <- as.data.frame(r1, xy = TRUE, na.rm = TRUE)
    
    attr(df, "resxy") <- terra::res(r1)  # guarda tamaño del píxel (lon/lat en grados)
    return(df)
  }

nice_breaks <- function(x, n = 7) pretty(x, n = n)


plot_raster <- function(df, value_col, title, subtitle, legend_title, outfile_png, aoi_sf) 
  {
    # Si no especificas value_col, usa la primera columna que no sea x/y
    if (missing(value_col)) value_col <- setdiff(names(df), c("x","y"))[1]
    resxy <- attr(df, "resxy")

      if(grepl("temp", title)) {
        paleta <- list("inferno", 0.5, 1.0) 
      } else paleta <- list("turbo", 0.1, 1)
    # scales::rescale(x, to = to, from = c(min(x, na.rm = TRUE), 1.2)
    
    gg <- ggplot() +
      geom_tile(data = df, 
                aes(x = x, y = y, fill = .data[[value_col]]),
                width  = resxy[1],     # ancho del píxel en grados
                height = resxy[2]) +   # alto del píxel en grados
      geom_sf(data = aoi_sf, fill = NA, color = "grey20", linewidth = 0.2) +
      coord_sf(crs = sf::st_crs(4326),         # == CRS de los datos → NO reproyecta
               default_crs = sf::st_crs(4326), # para capas sin CRS explícito
               expand = FALSE,
               datum = NA) +                   # sin graticulado/base
      labs(title = title, subtitle = subtitle, fill = legend_title) +
      scale_fill_viridis_c(option = paleta[[1]], na.value = NA,
                           begin = paleta[[2]], end = paleta[[3]]) +
      theme_minimal(base_size = 11) +
      theme(panel.grid = element_blank(),
            plot.title = element_text(size = 10, hjust = 0.5),
            plot.subtitle=element_text(size=8, hjust=0.5),
            axis.title.x = element_blank(),
            axis.title.y  = element_blank(),
            legend.text = element_text(size=6),
            legend.key.size = unit(0.5,"cm"))
     
     ggsave(outfile_png, gg, width = 8.5, height = 9.5, dpi = 300)
     saveRDS(gg, file = str_replace(outfile_png, pattern = "png$", replacement = "rds")) 
     invisible(gg)
     return(gg)
  }
# #------------------------
#   plot_raster(
#     df = df_dP,
#     value_col = names(df_dP)[3],
#     title = paste0("Δ Annual rainfall (%)"),
#     subtitle = sprintf("Scenario SSP%s vs 1970–2000; period %s",
#                        ssp, periodo),
#     legend_title = "Δ %",
#     outfile_png = file.path("_output", sprintf("latam_delta_Pann_pct_SSP%s_%s.png", 
#                             ssp, periodo)),
#     aoi_sf = aoi
#   )
# #------------------------

# Descarga: futuro CMIP6 y ensemble simple ----
get_future_stack <- function(var, ssp, model, periodo, res_wc) 
  {
    # var = "tn", "tm", o "prec"
    geodata::cmip6_world(var = var, res = res_wc, ssp = ssp,
                         model = model, time = periodo, path = "_data")
}


# Para cada SSP, construimos un ensemble promedio (promedio simple sobre 'models')
build_ensemble <- function(var, ssp) 
  {
    message(sprintf("Descargando futuro %s, SSP%s ...", var, ssp))
    stacks <- purrr::map(models, ~ get_future_stack(var, ssp, .x, periodo, res_wc))
    # Asegurar orden de capas y dimensiones
    # (cada stack trae 12 meses)
    s <- terra::sprc(stacks) |> terra::mosaic() # unión espacial segura
    # Promedio sobre modelos: apilamos por capa mensual (12 capas), luego mean
    # Para evitar problemas, calculamos mean por capa índice:
    layers <- seq_len(terra::nlyr(stacks[[1]]))
    mean_months <- lapply(layers, function(i) {
      terra::mean(terra::rast(purrr::map(stacks, ~ .x[[i]])))
    })
    terra::rast(mean_months)
  }


# Carpeta de trabajo / salidas ----
if (!dir.exists("_data"))
{
  dir.create("_data", showWarnings = FALSE)
  dir.create("_output", showWarnings = FALSE)
}

```

## Current state of climate change effects

Latin America is considered among the most **climate-vulnerable regions** (sharing risk with Africa, Small Island Developing States, South and Southeast Asia, Middle East and the Arctic), despite the relatively small contribution to global greenhouse gas emissions of all this regions.

The particular exposure of Latin America steam from a combination of structured factors. It is a region where climate hazards have a large probability to strike. This is a result of the geographical position. It spans the tropics, subtropics, and has an extensive coastlines, so it is exposed to hurricanes (Caribbean, Central America), El Niño/La Niña–driven droughts and floods (Pacific coast, Andes, Southern Cone), and heatwaves. Some sub-regions, like the tropical Andes, depend on glaciers as a source of water supply. There are ample sub-regions with low-lying coastal lands. River deltas, mangrove coastal forest and the Caribbean islands are clearly vulnerable to sea-level rise, storm surge and erosion.

The region is biodiversity rich. Five of the 17 megadiverse countries are located here: Brazil, Ecuador, Mexico, Peru and Venezuela. There are worldwide unique and fragile ecosystems located in the region. Some of them of global relevance: Amazon rainforets, Coral reefs, mangroves, páramo andino and tropical dry forest are widely recognized for its role in Holocene biosphere. The biocultural binding is strongly preserved in the region, making local livelihoods rely on these ecosystems for food, water, and raw materials.

But Latin America is also vulnerable because of socio-economic hardships. Inequity and poverty are widespread and pervasive, making difficult for people to cope with disasters. It is estimated that climate shocks push million into poverty or food insecurity. Cities grow through pattern of *informality*. This means that settlements are common to spread into floodplain, hillside or heat-prone territories. Related to all these, is the heavy economic and livelihood dependency on agriculture which makes food climate-dependent and spreads the human impact to wildlife and fisheries.

Therefore, Latin America an the Caribbean can be regarded to have a high physical exposure to climate change, has a global ecological importance. The human population in the region has economic dependence on climate-sensitive sectors, and social inequality provoke a state of “double vulnerability”: high sensibility to the impacts of climate change and limited adaptive capacity. A current state of perceived effects can be summarized as illustrated in @tbl-efectos. Expectations are that these effects will keep increasing.

In particular there is consensus that some tipping points could be near, if not already reached. For instance in the Amazon, if deforestation continues, large parts may shift to drier savanna-like systems, disrupting continental water cycles. Another wide consensus is that impacts are already visible and accelerating (heat, drought, floods, crop losses, zoonotic disease spread are increasing).

| Concept | Identified effect |
|-----------------------|-------------------------------------------------|
| **Climate patterns** | Average temperatures have risen; heatwaves are more frequent. Rainfall patterns are shifting: some areas are drier (more droughts), others have heavier rains. |
| **Extreme events** | Stronger and more frequent floods, droughts, wildfires, hurricanes, and intense storms. 2024 saw major hurricanes, floods, and droughts across the region. |
| **Water resources** | Andean glaciers are retreating rapidly, reducing dry-season water supply for drinking, irrigation, and hydropower. |
| **Ecosystems** | Amazon deforestation disrupts rainfall cycles and “flying rivers” (moisture flows). Coastal ecosystems (mangroves, reefs) suffer from sea-level rise and acidification. Species loss and species range shifts are already occurring. |
| **Food security** | Yields are falling in some crops due to heat and rainfall variability. Fisheries are stressed by ocean warming and acidification. Extreme events increase hunger risk in vulnerable populations. |
| **Zoonosis** | More mosquito-borne diseases (dengue, zika, chikungunya). Heatwaves raise mortality and stress health systems. Many areas lack infrastructure to cope with heat or disasters. |

: Some effects of climate change in Latin America that have been already detected {#tbl-efectos}

The patterns of climate change in terms of average temperature and rainfall are shown in @fig-mapas and @fig-mapas2. The base line values were calculated for the period 1970-2000. A projection to near future labeled as "2030" was calculated using data for the period 2021-2040.

```{r}
#| label: mapas
#| echo: false
#| message: false
#| warning: false

# Área de interés: América Latina ----
# Subregiones de Natural Earth: "South America", "Central America", "Caribbean"

ne <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf")
 
aoi <- ne |> 
  filter(subregion %in% c("South America", "Central America", "Caribbean") |
                          name %in% c("Mexico"))  |> 
  select(name, subregion)


# Parámetros de usuario ----
# Resolución: 10 minutos (~18 km al ecuador). Puedes cambiar a 2.5  (2.5 minutos) si tu RAM lo permite.
res_wc <- 10

# Escenarios y modelos CMIP6 (puedes ampliar esta lista)
# los distintos modelos serán promediados para cada escenario
ssps    <- c("245", "585")  # SSP2-4.5, SSP5-8.5   
models  <- c("BCC-CSM2-MR", "MIROC6", "CNRM-CM6-1")  # ejemplo de 3 modelos
periodo <- "2021-2040"      # "near term" ~ 2030

# Descarga: baseline (1970–2000) ----
# tavg: 12 capas mensuales (°C); prec: 12 capas mensuales (mm)

# if (length(list.files("_data/climate/wc2.1_10m/"))==0)
#{
  message("Descargando WorldClim v2.1 baseline...")
  tavg_base <- geodata::worldclim_global(var = "tavg", res = res_wc, path = "_data")
  prec_base <- geodata::worldclim_global(var = "prec", res = res_wc, path = "_data")
#}

  
# WorldClim tavg suele estar en °C (algunas versiones antiguas en 0.1 °C). Por si acaso:
if (max(terra::global(tavg_base, "max", na.rm = TRUE)) > 90) {
  tavg_base <- tavg_base / 10
}


# Agregados anuales baseline
tmean_base_ann <- terra::app(tavg_base, fun = mean)     # °C
prec_base_ann  <- terra::app(prec_base, fun = sum)      # mm/año 



# --- Futuro: construir ensemble anual por modelo y 
# promediar temperaturas max y min  por SSP---
# Construye ensambles 
per_ssp_tann <- list() 
per_ssp_pann <- list()
for (ssp in ssps) {
  tmin_f <- build_ensemble("tmin", ssp)
  tmax_f <- build_ensemble("tmax", ssp)
  pr_f   <- build_ensemble("prec", ssp)

  # tavg mensual como promedio capa a capa
  tavg_f <- (tmin_f + tmax_f) / 2

  # anual por píxel
  per_ssp_tann[[ssp]] <- terra::app(tavg_f, mean)  # °C
  per_ssp_pann[[ssp]] <- terra::app(pr_f,   sum)   # mm/año
}

tmean_fut_ann <- terra::mean(terra::rast(per_ssp_tann))  # ensemble simple
prec_fut_ann  <- terra::mean(terra::rast(per_ssp_pann))


mapas <-  list()
for (ssp in ssps) {
  message(sprintf("Calculando deltas para SSP%s ...", ssp))
  dT <- per_ssp_tann[[ssp]] - tmean_base_ann
  dP <- (per_ssp_pann[[ssp]] - prec_base_ann) / prec_base_ann * 100

  # Exportar GeoTIFF de datos
  terra::writeRaster(dT, filename = file.path("_output",
                     sprintf("latam_delta_Tmean_SSP%s_%s.tif", ssp, periodo)),
                     overwrite = TRUE)
  terra::writeRaster(dP, filename = file.path("_output", 
                     sprintf("latam_delta_Pann_pct_SSP%s_%s.tif", ssp, periodo)),
                     overwrite = TRUE)

  # DataFrames para mapas
  df_dT <- to_df(dT, aoi)
  df_dP <- to_df(dP, aoi)

  # Mapas PNG
mapa1 <- plot_raster(
    df = df_dT,
    value_col = names(df_dT)[3],
    title = paste0("Δ Mean annual temperature(°C)"),
    subtitle = sprintf("Scenario SSP%s vs 1970–2000; period %s",
                       ssp, periodo),
    legend_title = "Δ °C",
    outfile_png = file.path("_output", sprintf("latam_delta_Tmean_SSP%s_%s.png", 
                            ssp, periodo)),
    aoi_sf = aoi
  )

mapa2 <- plot_raster(
    df = df_dP,
    value_col = names(df_dP)[3],
    title = paste0("Δ Annual rainfall (%)"),
    subtitle = sprintf("Scenario SSP%s vs 1970–2000; period %s",
                       ssp, periodo),
    legend_title = "Δ %",
    outfile_png = file.path("_output", sprintf("latam_delta_Pann_pct_SSP%s_%s.png", 
                            ssp, periodo)),
    aoi_sf = aoi
  )
  mapas[[ssp]] <- list(mapa1, mapa2)
}

# Mapas de baseline (opcional): útil para contexto ----
df_Tbase <- to_df(tmean_base_ann, aoi)
df_Pbase <- to_df(prec_base_ann,  aoi)

g1 <- plot_raster(
  df = df_Tbase,
  value_col = names(df_Tbase)[3],
  title = "Mean annual temperature (°C) - baseline",
  subtitle = "WorldClim v2.1 1970–2000",
  legend_title = "°C",
  outfile_png = file.path("_output", "latam_Tmean_baseline_1970_2000.png"),
  aoi_sf = aoi
)

g2 <- plot_raster(
  df = df_Pbase,
  value_col = names(df_Pbase)[3],
  title = "Annual rainfall (mm) - baseline",
  subtitle = "WorldClim v2.1 1970–2000",
  legend_title = "mm/año",
  outfile_png = file.path("_output", "latam_Pann_baseline_1970_2000.png"),
  aoi_sf = aoi
)

message("Listo. Revisa la carpeta '_output' para los GeoTIFF y PNG.")

# mapas[[]]
#caption = paste0("Source: WorldClim v2.1 (baseline 1970–2000; ", 
#                            "CMIP6 2021–2040), Natural Earth")
```

```{r}
#| label: fig-mapas
#| fig-cap: Patterns of temperature in Latin America
#| echo: false



# g1  | mapas[["245"]][[1]] / mapas[["585"]][[1]]
g1  +  mapas[["585"]][[1]]



```


```{r}
#| label: fig-mapas2
#| fig-cap: Patterns of  rainfall in Latin America
#| echo: false

# g2 |  mapas[["245"]][[2]] / mapas[["585"]][[2]]
g2 + mapas[["585"]][[2]]

```


We are framing Latin America vulnerability to climate change using the **IUCN ecosystem typology** for climate action. In particular, IUCN has developed the *Ecosystem Red List & Key Biodiversity Areas (KBA)*, which are tools that help the identification of ecosystems at risk of collapse and areas of high conservation value. This assessment has gathered evidence hr support the view that in Latin America an the Caribbean region particular attention should be directed to Amazon, mangroves, páramos, tropical dry forests and coral reefs. A consensus perspective is emerging suggesting that strong commitments are required to protect–restore–manage habitats, reduce direct climate hazards, and address indirect human–climate synergies (@tbl-drivers).  

{{< include _drivers-cc-mammal-loss.qmd >}}


```{r}
#| label: tbl-drivers
#| tbl-cap: Main drivers of mammal biodiversity loss under current human climate change forcing.
#| echo: false

ft
```



